Psychological Needs During Intergroup Contact

### 2021-11-22 manuscript.tex
situational core need intro

\section{Psychological Needs in Intergroup Relations}
Although considerations of psychological needs and motivations have remained markedly absent within the intergroup contact literature, the concepts are in no way foreign to the general topic of intergroup relations. On the most general level, research on intergroup conflicts has, for example, shown that addressing the group-specific needs predicts willingness to reconcile with the other group \citep[][]{Shnabel2008}. For the case of refugee migrants in their relationship with a majority group, a recent study found that identity needs will buffer against discrimination experiences and will protect personal health \citep[][]{Celebi2017}.

Additionally, within the field of interpersonal relations, we find first evidence that the fulfillment of fundamental psychological needs might predict perceived interaction quality in non-intergroup interactions \citep[][]{Downie2008}. And even within the field of intergroup contact the idea of need fulfillment as a psychological mechanism is not entirely novel. \citet{Dovidio2017} in their narrative review have suggested that understanding psychological needs might be essential to constructive intergroup contact. And recently first empirical research has even found psychological need fulfillment during intergroup contacts to predict support for social change  \citep[][]{Hassler2021}. There is thus initial evidence to suggest that need fulfillment during intergroup contact might offer a promising psychological mechanism to understand perceived interaction quality and positive outgroup attitudes.

@@

Based on this we formulated four main hypotheses:
\begin{enumerate}
    \item 
    \item
    \item
\end{enumerate}

>>>>

\section{Psychological Needs in Intergroup Relations}
Although considerations of psychological needs and motivations have remained markedly absent within the intergroup contact literature, the concepts are in no way foreign to the general topic of intergroup relations. On the most general level, research on intergroup conflicts has, for example, shown that addressing the group-specific needs predicts willingness to reconcile with the other group \citep[][]{Shnabel2008}. For the case of refugee migrants in their relationship with a majority group, a recent study found that identity needs will buffer against discrimination experiences and will protect personal health \citep[][]{Celebi2017}. Additionally, within the field of interpersonal relations, we find first evidence that the fulfillment of fundamental psychological needs might predict perceived interaction quality in non-intergroup interactions \citep[][]{Downie2008}.

And even within the field of intergroup contact the idea of need fulfillment as a psychological mechanism is not entirely novel. \citet{Dovidio2017} in their narrative review have suggested that understanding psychological needs might be essential to constructive intergroup contact. And recently first empirical research has even found psychological need fulfillment during intergroup contacts to predict support for social change  \citep[][]{Hassler2021}. There is thus initial evidence to suggest that need fulfillment during intergroup contact might offer a promising psychological mechanism to understand perceived interaction quality and positive outgroup attitudes.

One reason why motivational considerations might have remained absent from the intergroup contact literature is that people might enter into a contact with a plethora of individual needs or goals that might be relevant to them in that particular situation. Researchers considering the motivational content would, thus, either test few hyper-specific needs that might not be transferable to other intergroup contexts or they may need to assess a broad range of fundamental psychological needs. To avoid this predicament we propose to start by using adaptive and responsive survey designs that allow a tailored approach based on the participants inputs \citep[e.g.,][]{Tourangeau2017}. In particular, we propose to ask the participants to report their main goal during the interaction in a short open-ended question (i.e., situation core need) and with reference to their own response the participants can then indicate how much this need was fulfilled during the interaction (i.e., need fulfillment). Such an adaptive approach allows us to test the fundamental psychological mechanism of whether situational psychological needs indeed predict perceived interaction quality and positive outgroup attitudes.

@@

Based on this we formulated four main hypotheses:
\begin{enumerate}
    \item Based on the most general understanding of the contact hypothesis, an increase in frequency and quality of contact should jointly account for changes in more favorable outgroup attitudes.
    \item Based on Allport’s optimal contact conditions, intergroup interactions with equal status, common goals, collaboration, and structural support should predict more favorable outgroup attitudes due to more positive interaction quality perceptions.
    \item Based on our proposal, intergroup interactions with higher situational core need fulfillment should predict more favorable outgroup attitudes due to more positive interaction quality perceptions.
    \item Based on our proposal, intergroup interactions with higher situational core need fulfillment should predict outgroup attitudes at least as well as Allport’s conditions.
\end{enumerate}
